Aspects of Dealing with Imperfect Data in Temporal Databases

### 2012-03-16 trunk/book/sources/fuzzyTemporalDatabases.tex
update from the task list.

\subsection{Handling Data Imperfections in Temporal Databases}

%The time when a package would arrive at its destination may be estimated and can eventually be researched, but it is typically not always precisely known. For such companies and in many other situations, information systems able to handle imperfection corresponding to certain temporal aspects of the objects modelled by the system are necessary. The transition between granularities~\cite{Lin97} is also sometimes considered as a source of imprecision. Therefore, some proposals consider granularity as the base of the temporal model~\cite{Cru97}.
%Some software applications deal with uncertainty related to time. e.g. a logistics company which transports packages from one destination to another. The time when the package would be at the end point may be estimated but not precisely known. These applications need to manage the uncertainty with respect to the temporal attributes of the objects stored in the database.

Data imperfections and techniques to represent them correctly in databases and queries are usually the focus of research in fuzzy databases. In fuzzy databases~\cite{Galindo2006}, uncertainty is usually expected to appear at storage level whereas imprecision is usually expected to appear in querying. Concerning temporal databases, there are several approaches to handle uncertainty in temporal data stored in a database.

Valid-time notions usually take the form of an interval. Some models consider probability~\cite{Dekhtyar2001} or possibility~\cite{Dubois89} distributions to describe the endpoints of a time interval. In this section, we will present a novel and currently researched technique to describe valid-time interval start- and endpoints using possibility distributions. In subsection \ref{subsec:representation-time-intervals} we will present an approach to modelling uncertainty in valid-time intervals. Finally, in subsection \ref{subsec:querying-time-intervals} the proposals for querying valid-time intervals in a flexible way are presented and discussed.%In this section we will explain the most interesting proposals that deal with possibility distributions in the valid-time interval. %Two are the main points developed within this section: the representation of uncertainty in the time interval and the flexible querying of ill-known time periods.

%In fuzzy databases~\cite{Galindo2006}, uncertainty is managed at storage level whereas imprecision is managed at querying level. In subsection \ref{subsec:representation-time-intervals} we will explain the main approach for the storage of uncertainty in the valid-time interval. Finally, in subsection \ref{subsec:querying-time-intervals} the proposals for querying in a flexible way valid-time intervals are analyzed.
\subsection{Handling Uncertainty in a Valid-time Relation}

\subsubsection{Ill-known Time Intervals}
\label{subsec:representation-time-intervals}
Uncertainty in a valid-time interval may be modelled by modelling the uncertainty concerning the exact start, respectively end point of the valid-time interval. The uncertainty concerning these points can by modelled by means of possibility distributions. It is of course also possible to use rough sets~\cite{Pawlak1995} for this modelling~\cite{Qia09}. The work presented in this section relies on possibilistic variables, which rely on possibility theory \cite{Dubois:Prade:1988:PossibilityTheory}.

A \emph{possibilistic variable} is defined as follows \cite{Pon11}.

\begin{svgraybox}
\vspace{-10pt}
\begin{definition}
A possibilistic variable $X$ over a universe $U$ is defined as a variable taking exactly one value in $U$, but for which this value is (partially) unknown. Its possibility distribution $\pi_X$ on $U$ models the available knowledge about the value that $X$ takes: for each $u\in U$, $\pi_X(u)$ represents the possibility that $X$ takes the value $u$. In this work, this possibility is interpreted as a measure of how plausible it is that $X$ takes the value $u$, given (partial) knowledge about the value $X$ takes.
\end{definition}
\vspace{-10pt}
\end{svgraybox}

The exact value a possibilistic variable takes, which is (partially) unknown, is called an \emph{ill-known value} in this work \cite{Dubois88b}.

When a possibilistic variable is defined on the powerset $\Pow(R)$ of some universe $R$, the unique value the variable takes will be a crisp set and its possibility distribution on the powerset $\Pow(R)$ will describe the possibility of each crisp subset of $R$ to be the value the variable takes. This exact value (a crisp set) the variable takes, is now called an \emph{ill-known set} \cite{Dubois88b}.

This work will deal with ill-known time intervals. These are ill-known intervals in time. Ill-known intervals are ill-known sets, defined and represented via a start and end point, which will be ill-known values. The elements of the set are the values between the start and end point. A closed ill-known interval with start point defined by possibilistic variable $X$ and end point by possibilistic variable $Y$ is noted here $\left[X, Y\right]$. The correspondences and transitions between the representations of ill-known sets, between the representations of ill-known intervals and between the representations of an ill-known set and an ill-known interval are part of the authors current research. Figure \ref{fig:interval} illustrates this.

@@

Several authors ~\cite{garrido2009} propose transformations in order to optimize the storage of such ill-known valid-time intervals, though recent research might seem to indicate some minor issues with respect to some of these transformations ~\cite{Pon11}. A comparison between the transformations in ~\cite{garrido2009} and the framework in~\cite{Pon11} is presented in~\cite{pon12}. Figure \ref{fig:convexhull} illustrates a transformation based on the `convex hull' approach REF.

%An ill-known time interval is usually represented by means of two possibility distributions one in each starting and ending points. Figure \ref{fig:interval} shows an ill-known valid-time interval. The starting and ending points are denoted by two fuzzy numbers, $X$ and $Y$, represented by means of two triangular possibility distributions (see the Appendix). For illustration, figure \ref{fig:convexhull} shows one of the transformations based in the convex hull and proposed in~\cite{garrido2009}.

\begin{figure}
\centering
\input{graphs/convexhull.tex}
\caption{Transformation based in the convex hull from the two ill-known points $X$ and $Y$. }
\label{fig:convexhull}
\end{figure}

In section \ref{subsec:querying-time-intervals}, a technique allowing the querying of fuzzy valid-time databases which contain ill-known valid-time intervals is presented and briefly explained. The technique relies on the framework for set evaluation presented in \cite{Pon11}.

\subsubsection{Querying Fuzzy Valid-Time Databases with Ill-Known Valid-Time Intervals}
\label{subsec:querying-time-intervals}

One of the main purposes of the existence of databases is to allow information retrieval. The standard query language for databases is SQL ~\cite{Mel93}. Several proposals to extend the SQL language for transaction-time databases~\cite{Sarda90}, valid-time databases~\cite{gad92} and bitemporal databases ~\cite{TSQL} exist and some authors have studied how to support temporal querying in standard SQL~\cite{Snodgrass98}.

In the following subsections, the query structure, the query evaluation method and the technique to meaningfully rank and aggregate the query results, proposed in this work, are presented.

@@

\subsubsection{Query Structure} TO BE ANOTHER PART
In this work, a query $\tilde{Q}$ is made up of two separate constructs of user query demands.

@@

\begin{svgraybox}
\vspace{-10pt}
\begin{definition}
\textbf{Query}
A query $\tilde{Q}$ in this work is given by:
\begin{equation}
\label{eq:query-specification}
\tilde{Q} = \left( Q^{time}, Q \right)
\end{equation}
Here $Q = \left \lbrace q_1,\ \cdots,\ q_n \right \rbrace$ contains all (possibly fuzzy) non-temporal user-defined query demands. These comprise all constraints unrelated to the valid-time indications in the queried relations. $Q^{time}$ denotes the temporal constraint specified by the user.
\end{definition}
\vspace{-10pt}
\end{svgraybox}
%%specify better this:
%\begin{svgraybox}
%\begin{definition}
%\label{def:query-constraint}
%\textbf{(query constraint)}
%$Q$ is the 
%A query constraint $q_a$ for an attribute $a$ with attribute domain $\mathcal{D}$, is a restriction in the subset of the values in the attribute domain:
%\begin{equation}
%\label{eq:query-constraint}
%q_a = \left \lbrace d \mid d \subseteq \mathcal{D} \right \rbrace
%\end{equation}
%\end{definition}

In this work, the query structure allows the user to specify a single temporal constraint.

\begin{svgraybox}
\vspace{-10pt}
\begin{definition}
\textbf{Temporal Constraint $Q^{time}$}
In this work, $Q^{time}$ is defined by:
\begin{equation}
Q^{time} = \left( I,AR \right)
\end{equation}
where $I$ denotes a crisp time interval and $AR$ is one of the Allen relations.
\end{definition}
\vspace{-10pt}
\end{svgraybox}

The interpretation is that, for a record with an ill-known valid-time interval $J$, the user demands that $I$ AR $J$ hold.

\subsubsection{Query evaluation} TO BE ANOTHER PART
\label{subsubsec:query-evaluation}
Query satisfaction in a fuzzy relational database is a matter of degree. Typically, the query evaluation results in a \emph{satisfaction degree} $s$, which is typically in the unit interval. $s \in \left[ 0,1,\right]$. Here, a satisfaction degree of $0$ denotes a total dissatisfaction while a degree of $1$ denotes a total satisfaction.

@@

\subsection{Bipolarity in Temporal Databases}
\label{subsubsec:bipolarity}
Humans express their preferences using both positive and negative statements, where positive statements express what is desired or acceptable and negative statements express what is undesired or unacceptable \cite{Billiet:Pons:Matthe:DeTre:Pons:2011:BipolarFuzzy}. This realization is interesting with regard to database querying, because sometimes a user does not exactly know his or her preferences or can't express them in only positive statements, but prefers to use negative statements to express what he or she dislikes or doesn't need. This introduces the need for \emph{bipolar querying}, a technique to model both positive and negative user preferences in a database query. Sometimes positive and negative preferences are clearly symmetric, making it possible to derive one from the other. For example, a person may define the concept of `tall' as `1.80 meters or higher'. The negative would then be the opposite: not tall would be `anyhing less than 1.80 meters'. However, in some cases, positive preferences can not be directly obtained from negative preferences or vice versa. E.g. when a person prefers to buy a black motorbike, this does not necessarily mean the person would totally reject a very dark blue motorbike. This phenomenon is called \emph{heterogeneous bipolarity}~\cite{Dubois2006},~\cite{Dubois2008}.

>>>>

\subsection{Data Imperfections in Temporal Databases}

%The time when a package would arrive at its destination may be estimated and can eventually be researched, but it is typically not always precisely known. For such companies and in many other situations, information systems able to handle imperfection corresponding to certain temporal aspects of the objects modelled by the system are necessary. The transition between granularities~\cite{Lin97} is also sometimes considered as a source of imprecision. Therefore, some proposals consider granularity as the base of the temporal model~\cite{Cru97}.
%Some software applications deal with uncertainty related to time. e.g., a logistics company which transports packages from one destination to another. The time when the package would be at the end point may be estimated but not precisely known. These applications need to manage the uncertainty with respect to the temporal attributes of the objects stored in the database.

Data and information imperfections and techniques to represent them correctly in databases and queries are usually the focus of research in fuzzy databases. Proposals from this field may present an approach based somehow on fuzzy set theory~\cite{Zadeh65} or possibility theory~\cite{Dubois:Prade:1988:PossibilityTheory}, although other theories support information imperfections too. Comparably, many proposals concerning the introduction of data imperfections or information imperfections in temporal databases present approaches based somehow on fuzzy set theory~\cite{garrido2009},~\cite{gal01},~\cite{nagypal2003},~\cite{Billiet:Pons:Matthe:DeTre:Pons:2011:BipolarFuzzy} or possibility theory~\cite{Dubois89},~\cite{Pon11},~\cite{129895}, although proposals based on other theories exist~\cite{Dyreson1998},~\cite{Dekhtyar2001},~\cite{Qia09}. As possibility theory is usually seen as a theory of confidence, aimed at dealing with uncertainty, in some proposals, possibility theory is used specifically to handle uncertainty in temporal information. In fuzzy databases~\cite{Galindo2006}, uncertainty is usually expected to appear in the database content, whereas other types of imperfection, notably imprecision, are usually expected to appear in querying.

Concerning temporal databases, there are several approaches to handle uncertainty in temporal data stored in a database. Many of these approaches concern several different types of time notions (VT, TT or DT), but most of these approaches focus somehow on valid-time~\cite{garrido2009},~\cite{gal01}.

In the following subsection, a novel approach to representing uncertainty concerning valid-time notions and a corresponding technique to query similar valid-time indications in a valid-time relation are proposed. The presented proposal is based on concepts introduced in~\cite{Dubois88b} and on the framework proposed in~\cite{Pon11}.

\subsection{Handling Uncertainty in a Valid-time Relation}
A valid-time indication usually takes the form of a time interval. Such a valid-time interval can be described (and stored in a valid-time relation record) using its boundaries (endpoints) or using one endpoint and the interval length. Usually, a valid-time interval is represented using its endpoints, which is also the approach adopted by the presented proposal.

Generally, the uncertainty concerning a set of values might be described by a possibility distribution on the powerset of which one of the elements can be the intended set~\cite{Dubois88b}. This representation, however, introduces some issues in practice or in practical applications. Therefore, in the presented proposal, possibility theory is used to model uncertainty, but only uncertainty concerning the exact values of the start and end point of a valid-time interval is considered and the uncertainty in both the start point and the end point are modelled using possibility distributions.

In fact, to model the uncertainty related to a valid-time interval using possibility theory, the presented proposal introduces so-called ill-known time intervals, relying on the concept of ill-known sets~\cite{Dubois88b}.

%Some models consider probability~\cite{Dekhtyar2001} or possibility~\cite{Dubois89} distributions to describe the endpoints of a time interval. In this section, we will present a novel and currently researched technique to describe valid-time interval start- and endpoints using possibility distributions. In subsection \ref{subsec:representation-time-intervals} we will present an approach to modelling uncertainty in valid-time intervals. Finally, in subsection \ref{subsec:querying-time-intervals} the proposals for querying valid-time intervals in a flexible way are presented and discussed.%In this section we will explain the most interesting proposals that deal with possibility distributions in the valid-time interval. %Two are the main points developed within this section: the representation of uncertainty in the time interval and the flexible querying of ill-known time periods.

%In fuzzy databases~\cite{Galindo2006}, uncertainty is managed at storage level whereas imprecision is managed at querying level. In subsection \ref{subsec:representation-time-intervals} we will explain the main approach for the storage of uncertainty in the valid-time interval. Finally, in subsection \ref{subsec:querying-time-intervals} the proposals for querying in a flexible way valid-time intervals are analyzed.

\subsubsection{Ill-known Time Intervals}
\label{subsec:representation-time-intervals}
%Uncertainty in a valid-time interval may be modelled by modelling the uncertainty concerning the exact start, respectively end point of the valid-time interval. The uncertainty concerning these points can by modelled by means of possibility distributions. It is of course also possible to use rough sets~\cite{Pawlak1995} for this modelling~\cite{Qia09}. The work presented in this section relies on possibilistic variables, which rely on possibility theory \cite{Dubois:Prade:1988:PossibilityTheory}.

To represent valid-time indications which might contain uncertainty, the presented proposal introduces the concept of ill-known valid-time intervals, which relies on the concept of ill-known sets~\cite{Dubois88b}. To correctly explain the concept of ill-known sets, the concept of possibilistic variables should be introduced first. In the presented proposal, the definition of possibilistic variables of~\cite{Pon11} is followed. In~\cite{Pon11}, a \emph{possibilistic variable} is defined as follows.

\begin{svgraybox}
\vspace{-10pt}
\begin{definition}
\textbf{Possibilistic variable}~\cite{Pon11}\\
A possibilistic variable $X$ over a universe $U$ is defined as a variable taking exactly one value in $U$, but for which this value is (partially) unknown. Its possibility distribution $\pi_X$ on $U$ models the available knowledge about the value that $X$ takes: for each $u\in U$, $\pi_X(u)$ represents the possibility that $X$ takes the value $u$. In the presented work, this possibility is interpreted as a measure of how plausible it is that $X$ takes the value $u$, given (partial) knowledge about the value $X$ takes.
\end{definition}
\vspace{-10pt}
\end{svgraybox}

Now, consider a set $R$, which contains single values (and not collections of values). When a possibilistic variable $X_{v}$ is defined on such a set $R$, the unique value $X_{v}$ takes, which is (partially) unknown, is called an \emph{ill-known value} in this work \cite{Dubois88b}.

When a possibilistic variable is defined on the powerset $\Pow(R)$ of some universe $R$, the unique value the variable takes will be a crisp set and its possibility distribution on the powerset $\Pow(R)$ will describe the possibility of each crisp subset of $R$ to be the value the variable takes. This value (a crisp set) the variable takes, which is (partially) unknown, is now called an \emph{ill-known set} \cite{Dubois88b}.

Finally, consider a set $R$, which contains single values (and not collections of values) and its powerset $\Pow(R)$. Now consider a subset $\Pow_{I}(R)$ of $\Pow(R)$ and let this subset contain every element of $\Pow(R)$ that is an interval, but no other elements. When a possibilistic variable $X_{i}$ is defined on the subset $\Pow_{I}(R)$ of the powerset $\Pow(R)$ of some set $R$, the unique value $X_{i}$ takes will be a crisp interval and the possibility distribution $\pi_{X_{i}}$ of $X_{i}$ will be a possibility distribution on $\Pow_{I}(R)$. This $\pi_{X_{i}}$ will define the possibility of each value of $\Pow_{I}(R)$ (a value of $\Pow_{I}(R)$ is a crisp interval which is a subset of $R$) being the value $X_{i}$ takes. This exact value (a crisp interval) the variable takes, which is (partially) unknown, is called an \emph{ill-known interval} here.

The presented proposal will deal with ill-known time intervals. Ill-known time intervals are ill-known intervals in a time domain. In the presented proposal, an ill-known time interval will be defined and represented via its start and end point, which will be ill-known values. The elements of the ill-known time interval are the values between its start and end point, including the start and end points themselves\footnote{The presented proposal only deals with closed ill-known time intervals. Dealing with halfopen or open ill-known intervals is part of the current research of the authors.}. It should be clear that this approach to representing ill-known time intervals differs from the approach based on a single possibility distribution on a set $\Pow_{I}(R)$ of a set $R$. These approaches have a different behavior and can be used to describe different ill-known time intervals. The correspondences, interactions and transitions between these different representations of ill-known intervals and their interpretations are part of the authors current research.

In the presented proposal, a closed ill-known time interval with start point defined by possibilistic variable $X$ and end point by possibilistic variable $Y$ is noted $\left[X, Y\right]$. Figure \ref{fig:interval} shows a closed ill-known time interval. %The correspondences and transitions between the representations of ill-known sets, between the representations of ill-known intervals and between the representations of an ill-known set and an ill-known interval are part of the authors current research. Figure \ref{fig:interval} illustrates this.

@@

Several authors work with concepts very similar to these ill-known time intervals and some of them~\cite{garrido2009} propose transformations of the describing functions in order to optimize the storage of such ill-known valid-time intervals, though recent research might seem to indicate some minor issues with respect to some of these transformations~\cite{Pon11}. A comparison between the transformations in~\cite{garrido2009} and the framework in~\cite{Pon11} is presented in~\cite{pon12}. Figure \ref{fig:convexhull} illustrates a transformation based on the `convex hull' approach~\cite{garrido2009}.

%An ill-known time interval is usually represented by means of two possibility distributions one in each starting and ending points. Figure \ref{fig:interval} shows an ill-known valid-time interval. The starting and ending points are denoted by two fuzzy numbers, $X$ and $Y$, represented by means of two triangular possibility distributions (see the Appendix). For illustration, figure \ref{fig:convexhull} shows one of the transformations based in the convex hull and proposed in~\cite{garrido2009}.
\begin{figure}
\centering
\input{graphs/convexhull.tex}
\caption{Transformation based in the convex hull from the two ill-known points $X$ and $Y$. }
\label{fig:convexhull}
\end{figure}

%In section \ref{subsec:querying-time-intervals}, a technique allowing the querying of fuzzy valid-time databases which contain ill-known valid-time intervals is presented and briefly explained. The technique relies on the framework for set evaluation presented in \cite{Pon11}.

In the presented proposal, ill-known valid-time intervals will be used to represent valid-time indications in a valid-time relation and to model the uncertainty these may contain.

To evaluate the temporal demands in queries issued by users to query a valid-time relation containing ill-known valid-time intervals, the presented proposal introduces a technique based on the concept of ill-known time constraints, which is based on the concept of ill-known constraints as presented in~\cite{Pon11}. Both concepts are treated in subsection \ref{iktc}.

Before ill-known time constraints can be introduced, another notion related to possibilistic variables shoud be paid attention to. In fact, a specific application of possibilistic variables is obtained when the set under consideration is the set of boolean values, denoted $\mathbb{B}$ = $\{T,F\}$, where $T$ denotes `true' and $F$ denotes `false' \cite{Pon11}. Indeed, any boolean proposition $p$ takes exactly one value in $\mathbb{B}$. If the knowledge about which value this proposition $p$ takes, is given by a possibility distribution $\pi_p$, then proposition $p$ can be seen as a possibilistic variable. In the presented proposal, the interest lies with the case where the proposition holds (denoted $p$ = $T$) and the possibility and necessity that $p$ = $T$ demand most attention. In the following sections, the following notations are used, based on previous notations:

\begin{align}
\text{Possibility that $p$ = $T$:} \hspace{50pt} & Pos(p) = \pi_p(T) \\
\text{Necessity that $p$ = $T$:} \hspace{50pt} & Nec(p) = 1-\pi_p(F) 
\end{align}

\subsubsection{\label{iktc}Ill-known Time Constraints}
The presented proposal contains a technique for evaluating user queries used to query a valid-time relation in which the valid-time indications are represented by ill-known time intervals. Part of this query evaluation technique relies on the concept of ill-known time constraints, which is based on the concept of ill-known constraints as presented in~\cite{Pon11}. These concepts are presented below. Following~\cite{Pon11}, an \emph{ill-known constraint} is defined as follows.

\begin{svgraybox}
\vspace{-10pt}
\begin{definition}
Given a universe $U$, an \emph{ill-known constraint} $C$ is specified by means of a binary relation $R \subseteq U^{2}$ and a fixed, ill-known value defined by its possibilistic variable $V$ on $U$, i.e.:
\begin{align}
C \triangleq (V,R) \nonumber
\end{align}
A set $A \subseteq U$ now satisfies this constraint $C$ if and only if:
\begin{align}
\forall a \in A : (V,a) \in R \nonumber
\end{align}
\end{definition}
\vspace{-10pt}
\end{svgraybox}

An example of an ill-known constraint is given by:
\begin{align}
C_{<} \triangleq (X, <) \nonumber
\end{align}
Some set $A$ then satisfies $C_{<}$ if 
\begin{align}
\forall a \in A : X <a \nonumber
\end{align}

The satisfaction of a constraint $C \triangleq (V,R)$ by a set $A$ is basically a Boolean matter (either the set satisfies the constraint or not) and can thus be seen as a boolean proposition, but due to the uncertainty inherent to the ill-known value $V$, it can be uncertain whether $C$ is satisfied by $A$ or not \cite{Pon11}. Based on the possibility distribution $\pi_{V}$ of $V$, the possibility and necessity that $A$ satisfies $C$ can be found. This proposition can thus be seen as a possibilistic variable on $\mathbb{B}$. The required possibility and necessity are calculated using the following formulas~\cite{Pon11}.

\begin{align}
Pos(A\text{ satisfies }C) & = \min_{a \in A}\left(\sup_{(w,a) \in R}\pi_{V}(w))\right) \label{eq:pos}\\
Nec(A\text{ satisfies }C) & = \min_{a \in A}\left(\inf_{(w,a) \notin R} 1-\pi_{V}(w)\right) \label{eq:nec}
\end{align}

Now, to calculate the possibility or necessity of a set $A$ satisfying multiple constraints, the $\min$ t-norm operator is used to express a conjunction of constraints. For example:

\begin{align}
Pos((A\text{ satisfies }C_{1}) \text{ and } (A\text{ satisfies }C_{2})) & = \nonumber\\
 \min_{a \in A}(Pos(A\text{ satisfies }C_{1}), Pos(A\text{ satisfies }C_{2})) \nonumber \\
Nec((A\text{ satisfies }C_{1}) \text{ and } (A\text{ satisfies }C_{2})) & = \nonumber\\
 \min_{a \in A}(Nec(A\text{ satisfies }C_{1}), Nec(A\text{ satisfies }C_{2})) \nonumber
\end{align}

Accordingly, the $\max$ t-conorm operator is used to express a disjunction of constraints. For example:

\begin{align}
Pos((A\text{ satisfies }C_{1}) \text{ or } (A\text{ satisfies }C_{2})) & = \nonumber\\
 \max_{a \in A}(Pos(A\text{ satisfies }C_{1}), Pos(A\text{ satisfies }C_{2})) \nonumber \\
Nec((A\text{ satisfies }C_{1}) \text{ or } (A\text{ satisfies }C_{2})) & = \nonumber\\
 \max_{a \in A}(Nec(A\text{ satisfies }C_{1}), Nec(A\text{ satisfies }C_{2})) \nonumber
\end{align}

%Negation?

In the presented proposal, \emph{ill-known time constraints} are considered, which are ill-known constraints of which the considered universe is a time domain.

In the next subsection, the core of the presented proposal is described.

\subsubsection{Querying Valid-time Relations containing Ill-known Valid-time Intervals}
\label{subsec:querying-time-intervals}

One of the main purposes of the existence of (relational) databases is to allow information retrieval. The standard query language for (relational) databases is SQL~\cite{Mel93}, but several proposals to extend the SQL language for transaction-time databases~\cite{Sarda90}, valid-time databases~\cite{gad92} and bitemporal databases~\cite{TSQL} exist and some authors have studied how to support temporal querying in standard SQL~\cite{Snodgrass98}.

As mentioned before, the presented proposal deals with querying a valid-time relation. In this subsection, the core of the presented proposal is described. First the particular structure of the relation is described, along with the nature and structure of its supposed contents. Next, the particular query structure is presented. Finally, the particular method for evaluating queries and for ranking the result records are presented.

%In the following subsections, the nature and structure of the relation content, the query structure, the query evaluation method and the technique to meaningfully rank and aggregate the query results, proposed in this work, are presented.

@@

\paragraph{\textbf{Relation Structure}}
In the presented proposal, a valid-time relation is considered, in which every record contains just one valid-time indication. This valid-time indication is represented by a closed ill-known time interval, to allow uncertainty in the valid-time information. As explained above, the ill-known time intervals used here will be defined and represented via their start and end points, which are ill-known values in the valid-time domain.

\paragraph{\textbf{Query Structure}}
In the presented proposal, a query consists of two separate constructs of user demands.

@@

\begin{svgraybox}
\vspace{-10pt}
\begin{definition}
\textbf{Query}\\
A query $\tilde{Q}$ is given by:
\begin{equation}
\label{eq:query-specification}
\tilde{Q} = \left( Q^{time}, Q \right)
\end{equation}

Here, $Q$ denotes the construct of all (possibly fuzzy) non-temporal user query demands. These comprise all constraints and demands unrelated to valid-time and thus unrelated to the valid-time indications in the queried relation. $Q^{time}$ denotes the temporal demand specified by the user.

%Here $Q = \left \lbrace q_1,\ \cdots,\ q_n \right \rbrace$ contains all (possibly fuzzy) non-temporal user-defined query demands. These comprise all constraints unrelated to the valid-time indications in the queried relations. $Q^{time}$ denotes the temporal constraint specified by the user.
\end{definition}
\vspace{-10pt}
\end{svgraybox}
%%specify better this:
%\begin{svgraybox}
%\begin{definition}
%\label{def:query-constraint}
%\textbf{(query constraint)}
%$Q$ is the 
%A query constraint $q_a$ for an attribute $a$ with attribute domain $\mathcal{D}$, is a restriction in the subset of the values in the attribute domain:
%\begin{equation}
%\label{eq:query-constraint}
%q_a = \left \lbrace d \mid d \subseteq \mathcal{D} \right \rbrace
%\end{equation}
%\end{definition}

The presented query structure allows the user to specify a single temporal demand, denoted by $Q^{time}$.

\begin{svgraybox}
\vspace{-10pt}
\begin{definition}
\textbf{Temporal demand}\\
A temporal demand $Q^{time}$ is defined by:
\begin{equation}
Q^{time} = \left( I,AR \right)
\end{equation}
Here, $I$ denotes a crisp time interval (which can be specified in any way required) and $AR$ denotes one of the Allen relations (cf. section~\ref{allen-crisp}).
\end{definition}
\vspace{-10pt}
\end{svgraybox}

The interpretation of such a temporal demand $Q^{time} = \left( I,AR \right)$ is that, for a record with an ill-known valid-time interval $J$, the user demands that $I$ AR $J$ holds.

\paragraph{\textbf{Query evaluation}}
\label{subsubsec:query-evaluation}
Query satisfaction in a fuzzy relational database is usually a matter of degree. Typically, the evaluation of the query demands for a record results in a \emph{satisfaction degree} $s$, which is typically in the unit interval, i.e. $s \in \left[ 0,1\right]$. This satisfaction degree then models the extent to which the record satisfies the query demands. As such, a satisfaction degree of $0$ indicates total dissatisfaction while a degree of $1$ indicates total satisfaction and every level of satisfaction between total satisfaction and total dissatisfaction is indicated by a satisfaction degre between $0$ and $1$.

@@

In the next subsection, some main concepts and issues concerning bipolarity in the context of temporal databases are presented and discussed.

\subsection{Bipolarity in Temporal Databases}
\label{subsubsec:bipolarity}
Humans express their preferences using both positive and negative statements, where positive statements express what is desired or acceptable and negative statements express what is undesired or unacceptable \cite{Billiet:Pons:Matthe:DeTre:Pons:2011:BipolarFuzzy}. This realization is interesting with regard to database querying, because sometimes a user does not exactly know his or her preferences or can't express them in only positive statements, but prefers to use negative statements to express what he or she dislikes or doesn't need. This introduces the need for \emph{bipolar querying}, a technique to model both positive and negative user preferences in a database query. Sometimes positive and negative preferences are clearly symmetric, making it possible to derive one from the other. For example, a person may define the concept of `tall' as `1.80 meters or higher'. The negative would then be the opposite: not tall would be `anything less than 1.80 meters'. However, in some cases, positive preferences can not be directly obtained from negative preferences or vice versa. e.g., when a person prefers to buy a black motorbike, this does not necessarily mean the person would totally reject a very dark blue motorbike. This phenomenon is called \emph{heterogeneous bipolarity}~\cite{Dubois2006},~\cite{Dubois2008}.
